---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov models.   
  The multilevel hidden Markov model (HMM) is a generalization of the well-known
  hidden Markov model, tailored to accommodate (intense) longitudinal data of 
  multiple individuals simultaneously. Using a multilevel framework, we allow 
  for heterogeneity in the model parameters (transition probability matrix and
  conditional distribution), while estimating one overall HMM. The model has a
  great potential of application in  many fields, such as the social sciences 
  and medicine. The model can be fitted on multivariate data with a categorical 
  distribution, and include individual level covariates (allowing for e.g.,
  group comparisons on model parameters). Parameters are estimated using 
  Bayesian estimation utilizing the forward-backward recursion within a hybrid 
  Metropolis within Gibbs sampler. The package also includes a function to 
  simulate data and a function to obtain the most likely hidden state sequence 
  for each individual using the Viterbi algorithm. 
  
  NOTE: this is a beta version of the package. The package is still (heavily) 
  under development, new functionalities will be added and functionalities will 
  be altered in the near future. 
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes")
```

## Usage

This is a basic example which shows you how to run the model using example data 
included with the package, and how to simulate data:

```{r example}
library(mHMMbayes)

##### Simple 2 state model
# specifying general model properties
m <- 2
n_dep <- 4
q_emis <- c(3, 2, 3, 2)

# specifying starting values
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05,
                          0.90, 0.05, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[1]), # vocalizing patient
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[2]), # looking patient
                  matrix(c(0.90, 0.05, 0.05,
                           0.05, 0.90, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[4])) # looking therapist

 # Run a model without covariates. 
 # Note that normally, a much higher number of iterations J would be used
 set.seed(23245)
 out_2st <- mHMM(s_data = nonverbal, 
              gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
              start_val = c(list(start_TM), start_EM),
              mcmc = list(J = 11, burn_in = 5))
 
out_2st
summary(out2_st)

# One can also plot the posterior densities for the transition and 
# emission probabilities:
plot(out1, component = "gamma")
plot(out1, component = "emiss", dep = 1)
plot(out1, component = "emiss", dep = 2)
plot(out1, component = "emiss", dep = 3)
plot(out1, component = "emiss", dep = 4)

# Run a model including a covariate 
# Here, the covariate (standardized CDI change) predicts the emission 
# distribution for each of the 4 dependent variables:
n_subj <- 10
xx <- rep(list(matrix(1, ncol = 1, nrow = n_subj)), (n_dep + 1))
for(i in 2:(n_dep + 1)){
 xx[[i]] <- cbind(xx[[i]], nonverbal_cov$std_CDI_change)
}
out_2st_c <- mHMM(s_data = nonverbal, xx = xx, 
                 gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
                 start_val = c(list(start_TM), start_EM),
                 mcmc = list(J = 11, burn_in = 5))

 
 ### Simulating data
 # simulating data for 10 subjects with each 100 observations
 T <- 100
 n <- 10
 m <- 3
 pr <- 4
 gamma <- matrix(c(0.8, 0.1, 0.1,
                   0.2, 0.7, 0.1,
                   0.2, 0.2, 0.6), ncol = m, byrow = TRUE)
 emiss_distr <- matrix(c(0.5, 0.5, 0.0, 0.0,
                         0.1, 0.1, 0.8, 0.0,
                         0.0, 0.0, 0.1, 0.9), nrow = m, ncol = pr, byrow = TRUE)
 set.seed(1253)
 data1 <- sim_mHMM(T = T, n = n, m = m, pr = pr, gamma = gamma, emiss_distr = emiss_distr,
                   var_gamma = 1, var_emiss = 1)


 # simulating subject specific transition probability matrices and emission distributions only
 T <- 0
 n <- 5
 m <- 3
 pr <- 4
 gamma <- matrix(c(0.8, 0.1, 0.1,
                   0.2, 0.7, 0.1,
                   0.2, 0.2, 0.6), ncol = m, byrow = TRUE)
 emiss_distr <- matrix(c(0.5, 0.5, 0.0, 0.0,
                         0.1, 0.1, 0.8, 0.0,
                         0.0, 0.0, 0.1, 0.9), nrow = m, ncol = pr, byrow = TRUE)
 set.seed(549801)
 data2 <- sim_mHMM(T = T, n = n, m = m, pr = pr, gamma = gamma, emiss_distr = emiss_distr,
                   var_gamma = 1, var_emiss = 1)
 data2

 set.seed(10893)
 data3 <- sim_mHMM(T = T, n = n, m = m, pr = pr, gamma = gamma, emiss_distr = emiss_distr,
                   var_gamma = .5, var_emiss = .5)
 data3


```
