---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<!-- badges: start -->
  [![R-CMD-check](https://github.com/emmekeaarts/mHMMbayes/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/emmekeaarts/mHMMbayes/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->

# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov
  models. The multilevel hidden Markov model (HMM) is a generalization of the
  well-known hidden Markov model, tailored to accommodate (intense) longitudinal
  data of multiple individuals simultaneously. Using a multilevel framework, we
  allow for heterogeneity in the model parameters (transition probability matrix
  and conditional distribution), while estimating one overall HMM. The model has
  a great potential of application in  many fields, such as the social sciences
  and medicine. The model can be fitted on multivariate data with categorical,
  continuous (i.e., normally distributed), or count (i.e., Poisson distributed)
  observations, and include individual level covariates (allowing for e.g.,
  group comparisons on model parameters). Parameters are estimated using
  Bayesian estimation utilizing the forward-backward recursion within a hybrid
  Metropolis within Gibbs sampler. Missing data (NA) in the dependent variables
  is accommodated assuming MAR. The package also includes various options for
  model visualization, a function to simulate data and a function to obtain the
  most likely hidden state sequence for each individual using the Viterbi
  algorithm.
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install the stable mHMMbayes package version directly from CRAN. Alternatively, you can install the GitHub version that includes the most recent package developments with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes")
```

## Usage
This is a basic example which shows you how to run the model using categorical example data 
included with the package. Below we include an example on continuous data and how to simulate data. For a more elaborate
introduction, see the vignette "tutorial-mhmm" accompanying the package.

```{r example1}
library(mHMMbayes)

##### Simple 2 state model
# specifying general model properties
m <- 2
n_dep <- 4
q_emiss <- c(3, 2, 3, 2)

# specifying starting values
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05,
                          0.90, 0.05, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[1]), # vocalizing patient
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[2]), # looking patient
                  matrix(c(0.90, 0.05, 0.05,
                           0.05, 0.90, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[4])) # looking therapist

 # Run a model without covariates. 
 # Note that normally, a much higher number of iterations J would be used
 set.seed(23245)
 out_2st <- mHMM(s_data = nonverbal, 
              gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
              start_val = c(list(start_TM), start_EM),
              mcmc = list(J = 11, burn_in = 5))
 
out_2st
summary(out_2st)

# Run a model including a covariate 
# Here, the covariate (standardized CDI change) predicts the emission 
# distribution for each of the 4 dependent variables:
n_subj <- 10
xx <- rep(list(matrix(1, ncol = 1, nrow = n_subj)), (n_dep + 1))
for(i in 2:(n_dep + 1)){
 xx[[i]] <- cbind(xx[[i]], nonverbal_cov$std_CDI_change)
}
out_2st_c <- mHMM(s_data = nonverbal, xx = xx, 
                 gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
                 start_val = c(list(start_TM), start_EM),
                 mcmc = list(J = 11, burn_in = 5))

```

This is a basic example which shows you how to run the model on data that includes continuous (i.e., normally distributed) observations, and how to simulate (this type of) data. 

```{r example2}
library(mHMMbayes)

###### Example on simulated data
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 50, 10,
                              100, 10,
                              150, 10), nrow = m, byrow = TRUE),
                    matrix(c(5, 2,
                             10, 5,
                             20, 3), nrow = m, byrow = TRUE))

set.seed(2327)
data_cont <- sim_mHMM(n_t = n_t, n = n, data_distr = 'continuous', gen = list(m = m, n_dep = n_dep),
                  gamma = gamma, emiss_distr = emiss_distr, var_gamma = .1, var_emiss = c(5^2, 0.2^2))

# Specify hyper-prior for the continuous emission distribution
manual_prior_emiss <- prior_emiss_cont(
                        gen = list(m = m, n_dep = n_dep),
                        emiss_mu0 = list(matrix(c(30, 70, 170), nrow = 1),
                                         matrix(c(7, 8, 18), nrow = 1)),
                        emiss_K0 = list(1, 1),
                        emiss_V =  list(rep(5^2, m), rep(0.5^2, m)),
                        emiss_nu = list(1, 1),
                        emiss_a0 = list(rep(1.5, m), rep(1, m)),
                        emiss_b0 = list(rep(20, m), rep(4, m)))

# Run the model on the simulated data:
set.seed(9834)
out_3st_cont_sim <- mHMM(s_data = data_cont$obs,
                         data_distr = 'continuous',
                         gen = list(m = m, n_dep = n_dep),
                         start_val = c(list(gamma), emiss_distr),
                         emiss_hyp_prior = manual_prior_emiss,
                          mcmc = list(J = 1000, burn_in = 200))

out_3st_cont_sim
summary(out_3st_cont_sim)

# obtaining the transition probability matrix gamma and the emission distribution 
# at the group and subject level 
obtain_gamma(out_3st_cont_sim, level = 'group')
obtain_emiss(out_3st_cont_sim, level = 'group')

obtain_gamma(out_3st_cont_sim, level = 'subject')
obtain_emiss(out_3st_cont_sim, level = 'subject')

# Inferring the most likely state at each point in time
inferred_states <- vit_mHMM(out_3st_cont_sim, data_cont$obs)
head(inferred_states)
```
