---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

*Note: this branch is dedicated to development of __using emission distributions of varying types within the same mHMM model__. When the development is complete (e.g., documentation is complete, functions thoroughly tested), the developments will be incorporated in the master branch, and eventually in the stable R CRAN version of the mHMM package.*

# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov 
  models. The multilevel hidden Markov model (HMM) is a generalization of the well-known
  hidden Markov model, tailored to accommodate (intense) longitudinal data of 
  multiple individuals simultaneously. Using a multilevel framework, we allow 
  for heterogeneity in the model parameters (transition probability matrix and
  conditional distribution), while estimating one overall HMM. The model has a
  great potential of application in  many fields, such as the social sciences 
  and medicine. The model can be fitted on multivariate data with a categorical 
  distribution, and include individual level covariates (allowing for e.g.,
  group comparisons on model parameters). Parameters are estimated using 
  Bayesian estimation utilizing the forward-backward recursion within a hybrid 
  Metropolis within Gibbs sampler. The package also includes various options for 
  model visualization, a function to simulate data and a function to obtain the 
  most likely hidden state sequence for each individual using the Viterbi algorithm. 
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes that incorporates the possibility of using emission distributions of varying types within the same mHMM model from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes@vary-emiss")
```

## Usage

This is a basic example which shows you how to run the model on data that includes categorical and continuous (i.e., normally distributed) observations, and how to simulate this type of data. 

```{r example}
library(mHMMbayes)

###### Example on simulated data
# simulating data with both categorical and continous dependent variables

# number of observations per subject, subjects, states and dependent variables
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 3

# distribution of each of the dependent variables
data_distr = c("categorical", "continuous", "continuous")

# number of categories for the categorical depenent variable
q_emiss <- c(4, 0, 0)

# transition probabilities between the states
gamma <- matrix(c(0.6, 0.3, 0.1,
                  0.2, 0.7, 0.1,
                  0.1, 0.4, 0.5), nrow = m, byrow = TRUE)

# emission distributions
                    # observation probabilities of each of the 4 categories of the first (categorical) dependent variable over the 3 states
emiss_distr <- list(matrix(c(0.80, 0.13, 0.05, 0.02,
                    0.08, 0.85, 0.05, 0.02,
                    0.05, 0.20, 0.40, 0.35), nrow = m, byrow = TRUE),
                    # mean and variance of second (continuous) dependend in each of the 3 states
                    matrix(c(6.0, 0.5,
                             5.0, 0.5,
                             3.0, 0.5), nrow = m, byrow = TRUE),
                    # mean and variance of third (continous) in each of the 3 states
                    matrix(c(4.0, 1.0,
                           5.0, 1.0,
                           6.5, 1.0), nrow = m, byrow = TRUE))

# amount of variance in the transition probabilities between subjects
var_gamma <- 0.01

# amount of variance in state dependent distributions of the three dependent variables between subjects
var_emiss <- c(0.01, 0.10, 0.10)

# simulate the data
data_vary <- sim_mHMM(n_t = n_t, n = n, data_distr = data_distr, m = m , n_dep = n_dep, q_emiss = q_emiss,
                  gamma = gamma, emiss_distr = emiss_distr, var_gamma = var_gamma, var_emiss = var_emiss)

# Specify hyper-prior for the continuous emission distribution
hyp_pr = emiss_cont_hyp_prior = list(
           emiss_mu0 = list(matrix(c(6,5,3), nrow = 1),
                            matrix(c(4,5,6), nrow = 1)),
           emiss_K0 = list(1,1),
           emiss_nu = list(1,1),
           emiss_V = list(rep(2, 3), rep(2, 3)),
           emiss_a0 = list(rep(1, 3), rep(1, 3)),
           emiss_b0 = list(rep(1, 3), rep(1, 3))
           )

# Run the model on the simulated data:
out_3st_vary_dep <- mHMM_vary(s_data = data_vary$obs,
                    gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss),
                    data_distr = data_distr,
                    start_val = c(list(gamma), emiss_distr),
                    emiss_cont_hyp_prior = hyp_pr,
                    mcmc = list(J = 11, burn_in = 5))

out_3st_vary_dep
summary(out_3st_vary_dep)

```
