---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

*Note: this branch is dedicated to development of a __singlelevel HMM using Bayesian estimation__. When the development is complete (e.g., documentation is complete, functions thoroughly tested), the developments will be incorporated in the master branch, and eventually in the stable R CRAN version of the mHMM package.*


# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov 
  models. The multilevel hidden Markov model (HMM) is a generalization of the well-known
  hidden Markov model, tailored to accommodate (intense) longitudinal data of 
  multiple individuals simultaneously. Using a multilevel framework, we allow 
  for heterogeneity in the model parameters (transition probability matrix and
  conditional distribution), while estimating one overall HMM. The model has a
  great potential of application in  many fields, such as the social sciences 
  and medicine. The model can be fitted on multivariate data with a categorical 
  distribution, and include individual level covariates (allowing for e.g.,
  group comparisons on model parameters). Parameters are estimated using 
  Bayesian estimation utilizing the forward-backward recursion within a hybrid 
  Metropolis within Gibbs sampler. The package also includes various options for 
  model visualization, a function to simulate data and a function to obtain the 
  most likely hidden state sequence for each individual using the Viterbi algorithm. 
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes  that incorporates the possibility of modelling singlelevel data  from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes@single-level-Bayes")
```

## Usage

This is a basic example which shows you how to run the model on singlelevel data, and how to simulate this type of data. 

```{r example}
library(mHMMbayes)

### first simulate data
# One subject with each n_t observations:
n <- 1
n_t <- 300

# specifying general model properties:
m <- 2
q_emiss <- 3
n_dep <- 1
gamma <- matrix(c(0.8, 0.2,
                  0.2, 0.8), ncol = m, byrow = TRUE)
emiss_distr <- list(matrix(c(0.50, 0.45, 0.05,
                        0.05, 0.05, 0.90), nrow = m, ncol = q_emiss, byrow = TRUE))

# simulate data
set.seed(97834)
data1 <- sim_mHMM(n_t = n_t, n = n, m = m, q_emiss = q_emiss,
                  gamma = gamma, emiss_distr = emiss_distr)

### analysis of data
# specifying starting values
st_gamma <- matrix(c(0.7, 0.3,
                     0.3, 0.7), ncol = m, byrow = TRUE)
st_emiss_distr <- list(matrix(c(0.40, 0.40, 0.20,
                           0.20, 0.20, 0.60), nrow = m, ncol = q_emiss, byrow = TRUE))

# run the model
set.seed(8750)
out_2st_sim <- HMM(s_data = data1$obs,
                    gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss),
                    start_val = c(st_gamma, st_emiss_distr),
                    mcmc = list(J = 4000, burn_in = 200), show_progress = TRUE)

out_2st_sim
summary(out_2st_sim)

```
