---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

*Note: this branch is dedicated to development of __fitting a model with FIXED GAUSSIAN emission distributions__ using the `mHMM_f()` function, creating an object of class `mHMM_f`. In this branch, the usual function `mHMM()` is not included. All S3 methods are tailored to the `mHMM_f()` function.*


# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov
  models. The multilevel hidden Markov model (HMM) is a generalization of the
  well-known hidden Markov model, tailored to accommodate (intense) longitudinal
  data of multiple individuals simultaneously. Using a multilevel framework, we 
  allow for heterogeneity in the model parameters (transition probability 
  matrix), while estimating one overall HMM. The model has a great potential of
  application in many fields, such as the social sciences and medicine.
  The model (in this specific branch) can be fitted on multivariate data with
  continuous (i.e., normally distributed) observations, and include individual 
  level covariates (allowing for e.g., group comparisons on model parameters). 
  Parameters are estimated using Bayesian estimation utilizing the 
  forward-backward recursion within a hybrid Metropolis within Gibbs sampler. 
  Missing data (NA) in the dependent variables is accommodated assuming MAR. The
  package also includes various options for model visualization, a
  function to simulate data and a function to obtain the most likely
  hidden state sequence for each individual using the Viterbi algorithm.
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes that accommodates fixed over subjects Gaussian emission distributions from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes@fixed-emiss")
```

## Usage
This is a basic example which shows you how to run the model using continuous data and how to simulate data. 


```{r example2}
library(mHMMbayes)

###### Example on simulated data
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 50, 10,
                              100, 10,
                              150, 10), nrow = m, byrow = TRUE),
                    matrix(c(5, 2,
                             10, 5,
                             20, 3), nrow = m, byrow = TRUE))

set.seed(2327)
data_cont <- sim_mHMM_f(n_t = n_t, n = n, data_distr = 'continuous', gen = list(m = m, n_dep = n_dep),
                  gamma = gamma, emiss_distr = emiss_distr, var_gamma = .1)

# Specify hyper-prior for the continuous emission distribution
manual_prior_emiss <- prior_emiss_cont_f(
                        gen = list(m = m, n_dep = n_dep),
                        emiss_mu0 = list(matrix(c(30, 70, 170), nrow = 1),
                                         matrix(c(7, 8, 18), nrow = 1)),
                        emiss_K0 = list(1, 1),
                        emiss_V =  list(rep(5^2, m), rep(0.5^2, m)),
                        emiss_nu = list(1, 1))

# Run the model on the simulated data:
set.seed(9834)
out_3st_cont_sim <- mHMM_f(s_data = data_cont$obs,
                         data_distr = 'continuous',
                         gen = list(m = m, n_dep = n_dep),
                         start_val = c(list(gamma), emiss_distr),
                         emiss_hyp_prior = manual_prior_emiss,
                          mcmc = list(J = 1000, burn_in = 200))

out_3st_cont_sim
summary(out_3st_cont_sim)

# obtaining the transition probability matrix gamma and the emission distribution 
# at the group level 
obtain_gamma(out_3st_cont_sim, level = 'group')
obtain_emiss_f(out_3st_cont_sim)       # level is not specified here as only group level parameters are available. 

# and ONLY for the transition probabilities at the subject level
obtain_gamma(out_3st_cont_sim, level = 'subject')

# Inferring the most likely state at each point in time
inferred_states <- vit_mHMM_f(out_3st_cont_sim, data_cont$obs)
head(inferred_states)
```
