---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

*Note: In contrast to the current CRAN version of mHMMbayes, this GitHub version allows for __using continuous observations (i.e., normally distributed)__ and missing data (NA) within the mHMM model. When this new version is even more thoroughly tested, the developments will be incorporated in the stable R CRAN version of the mHMM package.*


# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov
  models. The multilevel hidden Markov model (HMM) is a generalization of the
  well-known hidden Markov model, tailored to accommodate (intense) longitudinal
  data of multiple individuals simultaneously. Using a multilevel framework, we
  allow for heterogeneity in the model parameters (transition probability matrix
  and conditional distribution), while estimating one overall HMM. The model has
  a great potential of application in  many fields, such as the social sciences
  and medicine. The model can be fitted on multivariate data with categorical
  and/or continuous (i.e., normally distributed) observations, and include
  individual level covariates (allowing for e.g., group comparisons on model
  parameters). Parameters are estimated using Bayesian estimation utilizing the
  forward-backward recursion within a hybrid Metropolis within Gibbs sampler. 
  Missing data (NA) in the dependent variables is accommodated assuming MAR. 
  The package also includes various options for model visualization, a function
  to simulate data and a function to obtain the most likely hidden state
  sequence for each individual using the Viterbi algorithm.
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes that incorporates the possibility of modelling data with continuous observations from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes")
```

## Usage
This is a basic example which shows you how to run the model using categorical example data 
included with the package. Below we include an example on continuous data and how to simulate data. For a more elaborate
introduction, see the vignette "tutorial-mhmm" accompanying the package.

```{r example}
library(mHMMbayes)

##### Simple 2 state model
# specifying general model properties
m <- 2
n_dep <- 4
q_emiss <- c(3, 2, 3, 2)

# specifying starting values
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05,
                          0.90, 0.05, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[1]), # vocalizing patient
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[2]), # looking patient
                  matrix(c(0.90, 0.05, 0.05,
                           0.05, 0.90, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                  matrix(c(0.1, 0.9,
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[4])) # looking therapist

 # Run a model without covariates. 
 # Note that normally, a much higher number of iterations J would be used
 set.seed(23245)
 out_2st <- mHMM(s_data = nonverbal, 
              gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
              start_val = c(list(start_TM), start_EM),
              mcmc = list(J = 11, burn_in = 5))
 
out_2st
summary(out_2st)

# Run a model including a covariate 
# Here, the covariate (standardized CDI change) predicts the emission 
# distribution for each of the 4 dependent variables:
n_subj <- 10
xx <- rep(list(matrix(1, ncol = 1, nrow = n_subj)), (n_dep + 1))
for(i in 2:(n_dep + 1)){
 xx[[i]] <- cbind(xx[[i]], nonverbal_cov$std_CDI_change)
}
out_2st_c <- mHMM(s_data = nonverbal, xx = xx, 
                 gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
                 start_val = c(list(start_TM), start_EM),
                 mcmc = list(J = 11, burn_in = 5))

```

This is a basic example which shows you how to run the model on data that includes continuous (i.e., normally distributed) observations, and how to simulate (this type of) data. 

```{r example}
library(mHMMbayes)

###### Example on simulated data
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 5, 1,
                              10, 1,
                              15, 1), nrow = m, byrow = TRUE),
                    matrix(c(0.5, 0.1,
                             1.0, 0.2,
                             2.0, 0.1), nrow = m, byrow = TRUE))

set.seed(2327)
data_cont <- sim_mHMM(n_t = n_t, n = n, m = m, n_dep = n_dep, data_distr = 'continuous',
                  gamma = gamma, emiss_distr = emiss_distr, var_gamma = .1, var_emiss = c(.5, 0.01))

# Specify hyper-prior for the continuous emission distribution
hyp_pr <- list(
               emiss_mu0 = list(matrix(c(3,7,17), nrow = 1), matrix(c(0.7, 0.8, 1.8), nrow = 1)),
               emiss_K0  = list(1, 1),
               emiss_nu  = list(1, 1),
               emiss_V   = list(rep(2, m), rep(1, m)),
               emiss_a0  = list(rep(1, m), rep(1, m)),
               emiss_b0  = list(rep(1, m), rep(1, m)))

# Run the model on the simulated data:
set.seed(9834)
out_3st_cont_sim <- mHMM_cont(s_data = data_cont$obs,
                    gen = list(m = m, n_dep = n_dep),
                    start_val = c(list(gamma), emiss_distr),
                    emiss_hyp_prior = hyp_pr,
                    mcmc = list(J = 1000, burn_in = 200))

out_3st_cont_sim
summary(out_3st_cont_sim)

# obtaining the transition probability matrix gamma and the emission ditribution 
# at the group and subject level 
obtain_gamma(out_3st_cont_sim, level = 'group')
obtain_emiss(out_3st_cont_sim, level = 'subject')

# Inferring the most likely state at each point in time
inferred_states <- vit_mHMM_cont(out_3st_cont_sim, data_cont$obs)
head(inferred_states)
```
