---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

*Note: this branch is dedicated to development of __using continuous observations (i.e., normally distributed)__ within the mHMM model. This development is near complete. When the development is complete (e.g., documentation is complete, functions thoroughly tested), the developments will be incorporated in the master branch, and eventually in the stable R CRAN version of the mHMM package.*


# mHMMbayes

  With the \code{R} package mHMMbayes you can fit multilevel hidden Markov
  models. The multilevel hidden Markov model (HMM) is a generalization of the
  well-known hidden Markov model, tailored to accommodate (intense) longitudinal
  data of multiple individuals simultaneously. Using a multilevel framework, we
  allow for heterogeneity in the model parameters (transition probability matrix
  and conditional distribution), while estimating one overall HMM. The model has
  a great potential of application in  many fields, such as the social sciences
  and medicine. The model can be fitted on multivariate data with categorical
  and/or continuous (i.e., normally distributed) observations, and include
  individual level covariates (allowing for e.g., group comparisons on model
  parameters). Parameters are estimated using Bayesian estimation utilizing the
  forward-backward recursion within a hybrid Metropolis within Gibbs sampler.
  The package also includes various options for model visualization, a function
  to simulate data and a function to obtain the most likely hidden state
  sequence for each individual using the Viterbi algorithm.
  
  Please do not hesitate to contact me if you have any questions regarding the 
  package. 

## Installation

You can install mHMMbayes that incorporates the possibility of modelling data with continuous observations from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("emmekeaarts/mHMMbayes@continuous-emiss")
```

## Usage

This is a basic example which shows you how to run the model on data that includes continuous (i.e., normally distributed) observations, and how to simulate this type of data. 

```{r example}
library(mHMMbayes)

###### Example on simulated data
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 5, 1,
                              10, 1,
                              15, 1), nrow = m, byrow = TRUE),
                    matrix(c(0.5, 0.1,
                             1.0, 0.2,
                             2.0, 0.1), nrow = m, byrow = TRUE))

set.seed(2327)
data_cont <- sim_mHMM(n_t = n_t, n = n, m = m, n_dep = n_dep, data_distr = 'continuous',
                  gamma = gamma, emiss_distr = emiss_distr, var_gamma = .1, var_emiss = c(.5, 0.01))

# Specify hyper-prior for the continuous emission distribution
hyp_pr <- list(
               emiss_mu0 = list(matrix(c(3,7,17), nrow = 1), matrix(c(0.7, 0.8, 1.8), nrow = 1)),
               emiss_K0  = list(1, 1),
               emiss_nu  = list(1, 1),
               emiss_V   = list(rep(2, m), rep(1, m)),
               emiss_a0  = list(rep(1, m), rep(1, m)),
               emiss_b0  = list(rep(1, m), rep(1, m)))

# Run the model on the simulated data:
set.seed(9834)
out_3st_cont_sim <- mHMM_cont(s_data = data_cont$obs,
                    gen = list(m = m, n_dep = n_dep),
                    start_val = c(list(gamma), emiss_distr),
                    emiss_hyp_prior = hyp_pr,
                    mcmc = list(J = 1000, burn_in = 200))

out_3st_cont_sim
summary(out_3st_cont_sim)

# obtaining the transition probability matrix gamma and the emission ditribution 
# at the group and subject level 
obtain_gamma(out_3st_cont_sim, level = 'group')
obtain_emiss(out_3st_cont_sim, level = 'subject')

# Inferring the most likely state at each point in time
inferred_states <- vit_mHMM_cont(out_3st_cont_sim, data_cont$obs)
head(inferred_states)
```
