# Data {#sec-dat}
We illustrate using the `mHMMbayes` package using the embedded example data `nonverbal`. The data contains the nonverbal communication of *10 patient-therapist couples*, recorded for *15 minutes* at a frequency of *1 observation per second* (= 900 observations per couple).\
The following variables are contained in the dataset:

-   **`id`**: id variable of patient - therapist couple to distinguish which observation belongs to which couple.
-   **`p_verbalizing`**: verbalizing behavior of the patient, consisting of `1` = not verbalizing, `2` = verbalizing, `3` = back channeling.
-   **`p_looking`**: looking behavior of the patient, consisting of `1` = not looking at therapist, `2` = looking at therapist.
-   **`t_verbalizing`**: verbalizing behavior of the therapist, consisting of `1` = not verbalizing, `2` = verbalizing, `3` = back channeling.
-   **`t_looking`**: looking behavior of the therapist, consisting of `1` = not looking at patient, `2` = looking at patient. The top 6 rows of the dataset are provided below.

```{r load package and data, include = FALSE}
library(mHMMbayes)
nonverbal <- data.frame(nonverbal)
old_par <- graphics::par(no.readonly =TRUE)
```

<p>Below is a glimpse of how the data appears.</p>

<center>

```{r, echo=FALSE}
head(nonverbal)
```

</center>

When we plot the data of the first 5 minutes (= the first 300 observations) of the first couple, we get the following:

```{r plot observed data, fig.width = 7.2, fig.height = 3.5, echo = FALSE}
# set labels and colors for the observed behavioral categorical outcomes
library(RColorBrewer)
Voc_lab <- c("Not Speaking", "Speaking", "Back channeling")
Look_lab <-  c("Not looking", "Looking")
Voc_col <- c(brewer.pal(3,"PuBuGn")[c(1,3,2)])
Look_col <- c(brewer.pal(3,"YlOrRd")[-3])
cols = list(Voc_col, Look_col, Voc_col, Look_col)

time_s  <- seq(1,900)
couple1 <- cbind(nonverbal[nonverbal$id == 1,], time_s)

par(mar = c(4.3, 6.6, 2.1, 1.1))
plot(x = 1, xlim = c(0,300), ylim = c(0.5,6), type = "n", las = 1, xlab = "Time in minutes", xaxt = "n", yaxt = "n", ylab = "")
axis(2, at = seq(1,4), tick = FALSE, labels = c("P_vocalizing", "P_Looking", "T_vocalizing", "T_Looking"), las = 1)
axis(1, at = seq(0,300,60), tick = TRUE, las = 1, labels = FALSE)
axis(1, at = seq(0,300,60), tick = FALSE, las = 1, labels = seq(1,6,1))
abline(v = seq(0,300,60), col = "gray85")

for(j in 2:5){
  for(i in 1:max(nonverbal[,j])){
    points(x = couple1$time_s[1:300][couple1[1:300,j] == i], 
           y = rep(j-1, sum(couple1[1:300,j] == i)), 
           pch = "|", col = cols[[j-1]][i])
  }
}

legend("topright", bty = "n", fill = Voc_col, legend = Voc_lab)
legend("topleft", bty = "n", fill = Look_col, legend = Look_lab)

graphics::par(old_par)
```

We can, for example, observe that both the patient and the therapist are mainly looking at each other during the observed 5 minutes. During the first minute, the patient is primarily speaking. During the second minute, the therapists starts, after which the patient takes over while the therapist is back channeling.
