# Determining the most likely state sequence {#sec-sequence}


Given a well-fitting HMM, it may be of interest to determine the actual *sequence*, or order of succession, of hidden states that has most likely given rise to the sequence of outcomes as observed in a subject. One can either use local decoding, in which the probabilities of the hidden state sequence are obtained simultaneously with the model parameters estimates, or the well-known Viterbi algorithm [@viterbi1967; @forney1973]. In local decoding, the most likely state is determined separately at each time point $t$, in contrast to the Viterbi algorithm in which one determines the joint probability of the complete sequence of observations $O_{1:T}$ and the complete sequence of hidden states $S_{1:T}$.

In the package, local decoding can be achieved by saving the sampled hidden state sequence at each iteration of the Gibbs sampler, by setting the input variable `return_path = TRUE` for the function `mHMM`. This will result in very large output files, however. Global decoding can be performed by using the function `vit_mHMM`:

```{r using viterbi algorithm}
state_seq <- vit_mHMM(out_2st, s_data = nonverbal)
head(state_seq)
```

The function returns the hidden state sequence for each subject in a matrix, where each row represents a point in time and each column represents a subject. We can inspect the obtained hidden state sequence by for example plotting it together with the observed data. Below, the first 5 minutes of the first couple is plotted again, with the addition of the estimated state sequence:

```{r plotting observed data plus inferred states, fig.width = 7.2, fig.height = 4, echo = FALSE}
# set labels and colors for the observed behavioral categorical outcomes
Voc_lab <- c("Not Speaking", "Speaking", "Back channeling")
Look_lab <-  c("Not looking", "Looking")
Voc_col <- c(brewer.pal(3,"PuBuGn")[c(1,3,2)])
Look_col <- c(brewer.pal(3,"YlOrRd")[-3])
cols = list(Voc_col, Look_col, Voc_col, Look_col)

State_col <- c(rev(brewer.pal(3,"PiYG"))[-2])

time_s  <- seq(1,900)
couple1 <- cbind(nonverbal[nonverbal[,"id"] == 1,], time_s)

par(mar = c(4.3, 6.6, 2.1, 1.1))
plot(x = 1, xlim = c(0,300), ylim = c(-0.5,6), type = "n", las = 1, xlab = "Time in minutes", xaxt = "n", yaxt = "n", ylab = "")
axis(2, at = seq(0,4), tick = FALSE, labels = c("State", "P_vocalizing", "P_Looking", "T_vocalizing", "T_Looking"), las = 1)
axis(1, at = seq(0,300,60), tick = TRUE, las = 1, labels = FALSE)
axis(1, at = seq(0,300,60), tick = FALSE, las = 1, labels = seq(1,6,1))
abline(v = seq(0,300,60), col = "gray85")

for(j in 2:5){
  for(i in 1:max(nonverbal[,j])){
    points(x = couple1[,"time_s"][1:300][couple1[1:300,j] == i], 
           y = rep(j-1, sum(couple1[1:300,j] == i)), 
           pch = "|", col = cols[[j-1]][i])
  }
}
for(i in 1:2){
  points(x = couple1[,"time_s"][1:300][state_seq[1:300,1] == i], 
           y = rep(0, sum(state_seq[1:300,1] == i)), 
           pch = "|", col = State_col[i])
}

legend("topright", bty = "n", fill = c(Look_col, "white", Voc_col), legend = c(Look_lab, "",Voc_lab),
       ncol = 2, border = c(rep("black", 2), "white", rep("black", 3)))
legend("topleft", bty = "n", fill = State_col, legend = c("State 1", "State 2"))

graphics::par(old_par)

```
