# Model specification {#sec-modspec}

Let's proceed with fitting a simple 2-state multilevel model using the mHMM function. First, we need to set some general model properties and starting values:

```{r settings and load 2 state model, include = FALSE}
# specifying general model properties
m <- 2
n_dep <- 4
q_emiss <- c(3, 2, 3, 2)

# specifying starting values
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05, 
                          0.90, 0.05, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[1]), # vocalizing patient
                  matrix(c(0.1, 0.9, 
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[2]), # looking patient
                  matrix(c(0.90, 0.05, 0.05, 
                           0.05, 0.90, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                  matrix(c(0.1, 0.9, 
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[4])) # looking therapist

```

```{r show specifying 2 state model, eval = FALSE}

# Load the mHMMbayes package 
library(mHMMbayes)

# Specify general model properties:
m <- 2
n_dep <- 4
q_emiss <- c(3, 2, 3, 2)

# Specify starting values for transition matrix (TM) and emission probabilities (EM)
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05, 
                          0.90, 0.05, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[1]), # vocalizing patient
                  matrix(c(0.1, 0.9, 
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[2]), # looking patient
                  matrix(c(0.90, 0.05, 0.05, 
                           0.05, 0.90, 0.05), byrow = TRUE,
                         nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                  matrix(c(0.1, 0.9, 
                           0.1, 0.9), byrow = TRUE, nrow = m,
                         ncol = q_emiss[4])) # looking therapist
```

The first line of code loads the `mHMMbayes` package and the `nonverbal` data.\
Then, we specify the general model properties:

-   the number of states used is set by `m <- 2`.
-   the number of dependent variables in the dataset used to infer the hidden states is specified by `n_dep <- 4`.
-   the number of categorical outcomes for each of the dependent variables is specified by `q_emiss <- c(3, 2, 3, 2)`.

The subsequent lines of code specify the starting values for the transition probability matrix (`start_TM`) and the emission distributions (`start_EM`). Next, we will delve into more detail on how to specify these starting values.
